The Werewolf Among Us: Humans vs LLMs in Multi-Agent Games

### 2025-05-08 _tex/paper.tex
Built site for gh-pages

\title{The Werewolf Among Us: Humans vs LLMs in Multi-Agent Games}
\author{Bhavana Jonnalagadda \and Riley Jones}
\date{2025-05-07}
\begin{document}
\maketitle
\begin{abstract}
Abstract TODO
\end{abstract}

@@

Social deduction games like Werewolf offer a clear way to evaluate how
agents deceive, persuade, and reason in group settings
(\citeproc{ref-wikiwerewolf}{Wikipedia contributors 2024}). In these
games, players operate with limited information and hidden identities,
attempting to convince others while trying to uncover deception
themselves. These dynamics mirror real-world challenges involving trust,
negotiation, and manipulation. To explore how humans and large language
models (LLMs) navigate such scenarios, we analyzed two key datasets:
Werewolf Among Us (\citeproc{ref-laiWerewolfUsMultimodal2022}{Lai et al.
2022}), which features real human gameplay annotated with persuasion
strategies, and Werewolf Arena
(\citeproc{ref-bailisWerewolfArenaCase2024}{Bailis, Friedhoff, and Chen
2024}), a simulated environment in which LLM agents autonomously play
the game. While both studies demonstrate that Werewolf elicits rich,
strategic language, neither directly compares human and LLM behavior.
Our project fills this gap. We analyzed transcripts from both datasets,
aligning them by role, round, and persuasive strategy, and compared how
humans and LLMs lie, persuade, and detect deception. By annotating all
utterances using a consistent taxonomy of persuasive strategies, we
expose key differences and similarities in how synthetic agents and
humans handle adversarial group interactions.

@@

Recent research into multi-agent large language models (LLMs) has
explored their performance in various social deduction and collective
problem-solving contexts. Chi et al.~{[}Chi, Mao, and Tang
(\citeproc{ref-chiAMONGAGENTSEvaluatingLarge2024}{2024}){]} investigated
LLM behavior in the popular game Among Us, revealing capabilities in
understanding complex game dynamics and successfully navigating roles
involving deception and cooperation. Similarly, Du et al.~{[}Du,
Rajivan, and Gonzalez
(\citeproc{ref-duLargeLanguageModels2024}{2024}){]} examined collective
problem-solving scenarios, finding that LLM agent groups showed
increased complexity in their interactions, more frequent disagreements,
and generally positive exchanges compared to human groups. Piatti et
al.~{[}Piatti et al.
(\citeproc{ref-piattiCooperateCollapseEmergence2024}{2024}){]}, through
the GovSim environment, studied how AI societies managed collective
resources, demonstrating that LLM agents effectively balanced ethical
considerations, strategic planning, and negotiation, further supporting
the idea of their advanced cooperative and strategic capabilities.

Within the specific context of Werewolf, several studies have addressed
the use of LLMs to enhance gameplay. Xu et al.~{[}Xu et al.
(\citeproc{ref-xuLanguageAgentsReinforcement2024}{2024}){]} developed
LLM agents that leverage deductive reasoning and reinforcement learning
to optimize decision-making and gameplay strategy, outperforming
existing methods. Meanwhile, Bailis et al.~{[}Bailis, Friedhoff, and
Chen (\citeproc{ref-bailisWerewolfArenaCase2024}{2024}){]} introduced
the Werewolf Arena, a framework employed in our current study. However,
despite these advancements, previous research has not explicitly
examined or compared LLM-driven Werewolf gameplay to authentic human
interactions and strategies. Our paper specifically addresses this gap
by directly comparing human gameplay, sourced from the Werewolf Among Us
dataset, with simulated gameplay generated by LLMs, providing novel
insights into differences and similarities in persuasion strategies and
social reasoning between humans and AI.

@@

We used the \emph{Werewolf Among Us} dataset {[}Lai et al.
(\citeproc{ref-laiWerewolfUsMultimodal2022}{2022}){]}, containing
annotated dialogues from over 150 real games of One Night Werewolf and
Avalon. These games differ from classic Werewolf by having only one
round of discussion and voting, not eliminating players during gameplay,
and featuring specialized roles beyond Villager and Werewolf. The
dataset includes detailed annotations of persuasion strategies for each
utterance, such as accusations, defenses, and identity claims. Our
analysis specifically utilized textual transcriptions and strategy
annotations for direct comparison.

@@

The \emph{Werewolf Arena} dataset {[}Bailis, Friedhoff, and Chen
(\citeproc{ref-bailisWerewolfArenaCase2024}{2024}){]} comprises
simulated classic Werewolf games played by autonomous LLM agents. Unlike
one-round human games, these simulations include multiple rounds
alternating between night (secret actions) and day (open discussion).
Each agent receives a role (Villager, Werewolf, Seer, Doctor) and
interacts through tailored prompts generated via an LLM API.

@@

Agents participated in gameplay through a graphical user interface
(GUI), depicted in Figure~\ref{fig-wwa-gui}. The GUI displays the game
state, including player roles, actions, and current discussion rounds,
enabling monitoring of the gameplay progression.

@@

>>>>

\title{The Werewolf Among Us: Humans vs LLMs in Multi-Agent Games}
\author{Bhavana Jonnalagadda \and Riley Jones}
\date{2025-05-08}
\begin{document}
\maketitle
\begin{abstract}
Abstract TODO
\end{abstract}

@@

Social deduction games like Werewolf
(\citeproc{ref-wikiwerewolf}{Wikipedia contributors 2024}) offer a clear
way to evaluate how agents deceive, persuade, and reason in group
settings (\citeproc{ref-stepputtis2023longhorizon}{Stepputtis et al.
2023}). In these games, players operate with limited information and
hidden identities, attempting to convince others while trying to uncover
deception themselves. These dynamics mirror real-world challenges
involving trust, negotiation, and manipulation. To explore how humans
and large language models (LLMs) navigate such scenarios, we analyzed
two key datasets: Werewolf Among Us
(\citeproc{ref-laiWerewolfUsMultimodal2022}{Lai et al. 2022}), which
features real human gameplay annotated with persuasion strategies, and
Werewolf Arena (\citeproc{ref-bailisWerewolfArenaCase2024}{Bailis,
Friedhoff, and Chen 2024}), a simulated environment in which LLM agents
autonomously play the game. While both studies demonstrate that Werewolf
elicits rich, strategic language, neither directly compares human and
LLM behavior.

Our project fills this gap. We analyzed transcripts from both datasets,
aligning them by role, round, and persuasive strategy, and compared how
humans and LLMs lie, persuade, and detect deception. By annotating all
utterances using a consistent taxonomy of persuasive strategies, we
expose key differences and similarities in how synthetic agents and
humans handle adversarial group interactions.

@@

Recent research into multi-agent large language models (LLMs) has
explored their performance in various social deduction and collective
problem-solving contexts.
(\citeproc{ref-chiAMONGAGENTSEvaluatingLarge2024}{Chi, Mao, and Tang
2024}) investigated LLM behavior in the popular game Among Us, revealing
capabilities in understanding complex game dynamics and successfully
navigating roles involving deception and cooperation. Similarly, Du et
al. (\citeproc{ref-duLargeLanguageModels2024}{Du, Rajivan, and Gonzalez
2024}) examined collective problem-solving scenarios, finding that LLM
agent groups showed increased complexity in their interactions, more
frequent disagreements, and generally positive exchanges compared to
human groups.(\citeproc{ref-piattiCooperateCollapseEmergence2024}{Piatti
et al. 2024}), through the GovSim environment, studied how AI societies
managed collective resources, demonstrating that LLM agents effectively
balanced ethical considerations, strategic planning, and negotiation,
further supporting the idea of their advanced cooperative and strategic
capabilities.

Within the specific context of Werewolf, several studies have addressed
the use of LLMs to enhance gameplay.
(\citeproc{ref-xuLanguageAgentsReinforcement2024}{Xu et al. 2024})
developed LLM agents that leverage deductive reasoning and reinforcement
learning to optimize decision-making and gameplay strategy,
outperforming existing methods. Meanwhile,
(\citeproc{ref-bailisWerewolfArenaCase2024}{Bailis, Friedhoff, and Chen
2024}) introduced the Werewolf Arena, a framework employed in our
current study. However, despite these advancements, previous research
has not explicitly examined or compared LLM-driven Werewolf gameplay to
authentic human interactions and strategies.

@@

We used the \emph{Werewolf Among Us} dataset
(\citeproc{ref-laiWerewolfUsMultimodal2022}{Lai et al. 2022}),
containing annotated dialogues from over 150 real games of One Night
Werewolf and Avalon. These games differ from classic Werewolf by having
only one round of discussion and voting, not eliminating players during
gameplay, and featuring specialized roles beyond Villager and Werewolf.
The dataset includes detailed annotations of persuasion strategies for
each utterance, such as accusations, defenses, and identity claims. Our
analysis specifically used textual transcriptions and strategy
annotations for direct comparison.

@@

The \emph{Werewolf Arena} dataset
(\citeproc{ref-bailisWerewolfArenaCase2024}{Bailis, Friedhoff, and Chen
2024}) comprises simulated classic Werewolf games played by autonomous
LLM agents. Unlike one-round human games, these simulations include
multiple rounds alternating between night (secret actions) and day (open
discussion). Each agent receives a role (Villager, Werewolf, Seer,
Doctor) and interacts through tailored prompts generated via an LLM API.

Agents interact with the game interface via specialized prompts
reflecting their current role, memory state, and game context. The
prompts guide strategic interactions, influencing agent decisions in
voting, debating, and night actions. After generating dialogues through
the LLM API, we manually annotated these interactions using the
persuasion strategy categories from the human dataset.

@@

LLM agent gameplay can be viewed and debugged through a graphical user
interface (GUI), depicted in Figure~\ref{fig-wwa-gui}. The GUI displays
the game state, including player roles, actions, utterances,
inner-monologue and current discussion rounds, enabling monitoring of
the gameplay progression.

@@

\bibitem[\citeproctext]{ref-stepputtis2023longhorizon}
Stepputtis, Simon, Joseph Campbell, Yaqi Xie, Zhengyang Qi, Wenxin
Sharon Zhang, Ruiyi Wang, Sanketh Rangreji, Charles Michael Lewis, and
Katia P. Sycara. 2023. {``Long-Horizon Dialogue Understanding for Role
Identification in the Game of Avalon with Large Language Models.''} In
\emph{Proceedings of the 2023 Conference on Empirical Methods in Natural
Language Processing}. \url{https://openreview.net/forum?id=JKmsjKJ0Q8}.
